High-Dimensional Regularized Discriminant Analysis

### 2013-05-11 reports/paper-hdrda.Rnw
Minor edits before customizing paper for AoAS.

Microarrays have enabled researchers to measure gene-expression levels of
thousands of genes in a single experiment to identify a subset of genes and
their expression patterns that either characterize a particular cell state or
facilitate an accurate prediction of future cell state.  We focus on the
classification of tissues and other biological samples into one of $K$ disjoint,
predefined cancer types, where our goal is to ascertain the class membership of
a new tissue sample based on gene-expression microarray data. However,
microarray data typically consist of a large number of gene expressions $p$
relative to the total sample size $N$. Moreover, standard benchmark microarray
data sets usually consist of between $p = 10,000$ and $55,000$ probes and fewer
than $N = 100$ observations, in which case standard supervised-learning methods
exhibit poor classification performance because of the \emph{curse of
  dimensionality} \citep{Bellman:1961tn}. Furthermore, some well-known
classifiers, such as linear and quadratic discriminant analysis, are
incalculable when $p > N$ because the sample covariance matrices employed are
singular. As a result, many researchers, including \cite{Mai:2012bf},
\cite{Fan:2012iq}, \cite{Merchante:2012vk}, \cite{Clemmensen:2011kr}, and
\cite{Huang:2010ju}, have introduced restrictions on the population covariance
matrices in linear and quadratic discriminant analysis to reduce the number of
covariance parameters that must be estimated to ensure that the resulting
classifiers are calculable. For instance, \cite{Dudoit:2002ev} have modified the
linear and quadratic discriminant analysis classifiers by assuming that the
population covariance matrices are diagonal, resulting in classifiers that are
special cases of the Naive Bayes classifier \citep[Chapter 4]{Hastie:2008dt}.

@@

Here, we propose the high-dimensional regularized discriminant analysis
(\emph{HDRDA}) by first constructing a pooling matrix as a convex combination of
the covariance-matrix estimators employed in linear and quadratic discriminant
analysis. Given that the assumption of equal population covariance matrices for
linear discriminant analysis is often too stringent in practice
\citep{Clemmensen:2011kr}, our proposed classifier can relax the linearity
assumption by instead incorporating unequal covariance matrices while retaining
linear and quadratic discriminant analysis as special cases. We then establish
\emph{HDRDA} classifier by applying covariance-matrix regularization with a
parameterization that includes a large family of shrinkage methods, such as
those from \cite{Srivastava:2007ww}, \cite{Rao:1971ul}, and other methods
studied by \cite{Ramey:2013ji} and \cite{Xu:2009fl}. This regularization has the
effect of stabilizing the eigenvalues of the covariance-matrix estimators and
ensuring that the classifier is well-posed.

We show that our proposed classifier is invariant to any adjustments of the
eigenvalues corresponding to the null space of the pooled sample covariance
matrix. Applying a reasoning similar to \cite{Ye:2006tq}, we decompose the
decision rule of the \emph{HDRDA} classifier and show that the sum related to
the null space of the pooled sample covariance matrix can be discarded. Hence,
no gain in classification performance can be obtained by adjusting these zero
eigenvalues. This invariance property provides a substantial increase in
computational efficiency of model training and selection. In fact, we show that
the inverse and determinants can be reduced to calculations of $n_k \times n_k$
matrices, where $n_k$ is the sample size of the $k$th class $(k = 1, \ldots,
K)$. This reduction in calculation is clearly advantageous because $n_k$ is
often quite small in microarray data sets (e.g., 15--20 observations).

@@

The smallest eigenvalues and the directions associated with their eigenvectors
highly influence \eqref{eq:qda}. In fact, the eigenvalues of $\widehat{ \bm
  \Sigma}_k$ are well-known to be biased if $p \ge n_k$ such that the smallest
eigenvalues are underestimated \citep{Seber:2004uh}, Moreover, if $p > n_k$,
then rank$(\widehat{\bm \Sigma}_k) \le n_k$, which implies that at least $p -
n_k$ eigenvalues of $\widehat{\bm \Sigma}_k$ are 0. Furthermore, although more
feature information is available to discriminate among the $K$ classes, if $p >
n_k$, \eqref{eq:qda} is incalculable because $\widehat{\bm \Sigma}_k^{-1}$ does
not exist.

@@

\eqref{eq:hdrda-cov} into \eqref{eq:qda}, we define the high-dimensional
regularized discriminant analysis classifier

@@

Thus far, we have shown that our proposed classifier in \eqref{eq:hdrda} is
invariant to the term $\bm U_2$, thus yielding an equivalent classifier in
\eqref{eq:hdrda-decomposed} with a substantial reduction in the computational
complexity. Here, we demonstrate that the computational efficiency in
calculating the inverse and determinants of $\bm W_k$ can be further improved
via standard matrix operations.

@@

We evaluated the classification performance of each classifier by randomly
partitioning the data such that $2/3$ of the observations were allocated as
training data and the remaining $1/3$ of the observations were allocated as a
test data set. For each random partition, we applied the feature-selection
method from \cite{Dudoit:2002ev} to the training data to choose the $d$ features
having the largest ratio of their between-group to within-group sums of squares.
We then filtered the test data set with the corresponding $d$ features. After
training each classifier from the filtered training data, we classified the
filtered test data sets and computed the proportion of mislabeled test
observations to estimate the classification error rate for each
classifier. Repeating this process 1000 times, we computed the average of the
error-rate estimates for each classifier. We remark that our simulation design
avoids the selection bias discussed in \cite{Ambroise:2002fa}.

@@

For the microarray data sets, we computed error-rate estimates for the competing
classifiers after reducing the dimension of the data sets with feature selection
to $d = 100, 200, \ldots, 1000$, similar to the values of $d$ considered by
\cite{Xu:2009fl}. To compare misclassification rates between different values of
$d$, we applied each value of $d$ considered to the same random allocation of
training and test data. Hence, for a given random partition, the set of features
selected for a small value of $d$ is a proper subset of the features selected
for a larger value of $d$. We report the average of the error-rate estimates for
each value of $d$ in Figure \ref{fig:error-rates}. Approximate standard errors
were no greater than \Sexpr{max_se}.

@@

The classification performance of the the Dudoit, Pang, Tong, and Witten
classifiers for the Shipp and Singh data sets revealed the inability of these
methods to handle larger values of $d$. After closer examination of the
classification studies performed in the authors' original papers, we noticed
that classification accuracy estimates were reported only for small values of
$d$, thereby ignoring the inclusion of a larger number of variables. Hence, we
advise that practioners be aware of the necessity to filter a large number of
features before employing these methods, lest their classification performance
be degraded.

@@

In our simulation study, we compared our proposed \emph{HDRDA} classifier to
several recently proposed classifiers, including the Dudoit, Pang, Tong, and Guo
classifiers, each of which assumes independence of the features conditional on
class membership. Despite the rapid computational performance of diagonal
classifiers and the reduction in the number of parameters to estimate,
\cite{Mai:2012bf} and \cite{Fan:2012iq} have noted that diagonal classifiers can
often yield inferior classification performance to other classification
methods. We have shown that the \emph{HDRDA} classifier often yields superior
classification accuracy to the diagonal classifiers considered, confirming the
assertions of \cite{Mai:2012bf} and \cite{Fan:2012iq}. Furthermore, the
implementation of our classifier in the {\tt sparsediscrim} R package was
comparable to that of the diagonal classifiers in terms of computational
runtime. Moreover, when preliminary variable selection was applied to the data
in our classification study, the time to perform model selection with our
proposed classifier was nearly instantaneous. Hence, our proposed \emph{HDRDA}
classifier is competitive in terms of classification accuracy for modern
small-sample, high-dimensional data sets and requires substantially less
computing time than standard regularized classifiers, such as regularized
discriminant analysis proposed by \cite{Friedman:1989tm}.

>>>>

Microarrays have enabled researchers to measure gene-expression levels of
thousands of genes in a single experiment to identify a subset of genes and
their expression patterns that either characterize a particular cell state or
facilitate an accurate prediction of future cell state.  We focus on the
classification of tissues and other biological samples into one of $K$ disjoint,
predefined types, where our goal is to ascertain the class membership of
a new tissue sample based on gene-expression microarray data. However,
microarray data typically consist of a large number of gene expressions $p$
relative to the total sample size $N$. Moreover, standard benchmark microarray
data sets usually consist of between $p = 10,000$ and $55,000$ probes and fewer
than $N = 100$ observations, in which case standard supervised-learning methods
exhibit poor classification performance because of the \emph{curse of
  dimensionality} \citep{Bellman:1961tn}. Furthermore, some well-known
classifiers, such as linear and quadratic discriminant analysis, are
incalculable when $p > N$ because the sample covariance matrices employed are
singular. As a result, many researchers, including \cite{Mai:2012bf},
\cite{Fan:2012iq}, \cite{Merchante:2012vk}, \cite{Clemmensen:2011kr}, and
\cite{Huang:2010ju}, have introduced restrictions on the population covariance
matrices in linear and quadratic discriminant analysis to reduce the number of
covariance parameters that must be estimated to ensure that the resulting
classifiers are calculable. For instance, \cite{Dudoit:2002ev} have modified the
linear and quadratic discriminant analysis classifiers by assuming that the
population covariance matrices are diagonal, resulting in classifiers that are
special cases of the Naive Bayes classifier \citep[Chapter 4]{Hastie:2008dt}.

@@

Here, we propose the high-dimensional regularized discriminant analysis
(\emph{HDRDA}) by first constructing a pooling matrix as a convex combination of
the covariance-matrix estimators employed in linear and quadratic discriminant
analysis. Given that the assumption of equal population covariance matrices for
linear discriminant analysis is often too stringent in practice
\citep{Clemmensen:2011kr}, our proposed classifier can relax the linearity
assumption by instead incorporating unequal covariance matrices while retaining
linear and quadratic discriminant analysis as special cases. We then establish
the \emph{HDRDA} classifier by applying covariance-matrix regularization with a
parameterization that includes a large family of shrinkage methods, such as
those from \cite{Srivastava:2007ww}, \cite{Rao:1971ul}, and other methods
studied by \cite{Ramey:2013ji} and \cite{Xu:2009fl}. This regularization has the
effect of stabilizing the eigenvalues of the covariance-matrix estimators and
ensuring that the classifier is well-posed.

We show that our proposed classifier is invariant to any adjustments of the
eigenvalues corresponding to the null space of the pooled sample covariance
matrix. Applying a reasoning similar to \cite{Ye:2006tq}, we decompose the
decision rule of the \emph{HDRDA} classifier and show that the sum related to
the null space of the pooled sample covariance matrix can be discarded. Hence,
no gain in classification performance can be obtained by adjusting these zero
eigenvalues. This invariance property provides a substantial increase in
computational efficiency of model training and selection. In fact, we show that
the inverse and determinant can be reduced to calculations of $n_k \times n_k$
matrices, where $n_k$ is the sample size of the $k$th class $(k = 1, \ldots,
K)$. This reduction in calculation is clearly advantageous because $n_k$ is
often quite small in microarray data sets (e.g., 10--20 observations).

@@

The smallest eigenvalues of $\widehat{\bm \Sigma}_k$ and the directions
associated with their eigenvectors highly influence \eqref{eq:qda}. In fact, the
eigenvalues of $\widehat{ \bm \Sigma}_k$ are well-known to be biased if $p \ge
n_k$ such that the smallest eigenvalues are underestimated \citep{Seber:2004uh},
Moreover, if $p > n_k$, then rank$(\widehat{\bm \Sigma}_k) \le n_k$, which
implies that at least $p - n_k$ eigenvalues of $\widehat{\bm \Sigma}_k$ are
0. Furthermore, although more feature information is available to discriminate
among the $K$ classes, if $p > n_k$, \eqref{eq:qda} is incalculable because
$\widehat{\bm \Sigma}_k^{-1}$ does not exist.

@@

\eqref{eq:hdrda-cov} into \eqref{eq:qda}, we define the \emph{HDRDA} classifier

@@

Thus far, we have shown that our proposed classifier in \eqref{eq:hdrda} is
invariant to the term $\bm U_2$, thus yielding an equivalent classifier in
\eqref{eq:hdrda-decomposed} with a substantial reduction in the computational
complexity. Here, we demonstrate that the computational efficiency in
calculating the inverse and determinant of $\bm W_k$ can be further improved
via standard matrix operations.

@@

We evaluated the classification performance of each classifier by randomly
partitioning the data set under consideration such that $2/3$ of the
observations were allocated as training data and the remaining $1/3$ of the
observations were allocated as a test data set. For each random partition, we
applied the feature-selection method from \cite{Dudoit:2002ev} to the training
data to choose the $d$ features having the largest ratio of their between-group
to within-group sums of squares.  We then filtered the test data set with the
corresponding $d$ features. After training each classifier from the filtered
training data, we classified the filtered test data sets and computed the
proportion of mislabeled test observations to estimate the classification error
rate for each classifier. Repeating this process 1000 times, we computed the
average of the error-rate estimates for each classifier. We remark that our
simulation design avoids the selection bias discussed in \cite{Ambroise:2002fa}.

@@

For the microarray data sets, we computed error-rate estimates for the competing
classifiers after reducing the dimension of the data sets with feature selection
to $d = 100, 200, \ldots, 1000$, similar to the values of $d$ considered by
\cite{Xu:2009fl}. To compare misclassification rates between different values of
$d$, we applied each value of $d$ considered to the same random allocation of
training and test data. Hence, for a given random partition, the set of features
selected for a smaller value of $d$ is a proper subset of the features selected
for a larger value of $d$. We report the average of the error-rate estimates for
each value of $d$ in Figure \ref{fig:error-rates}. Approximate standard errors
were no greater than \Sexpr{max_se}.

@@

The classification performance of the the Dudoit, Pang, Tong, and Witten
classifiers for the Shipp and Singh data sets revealed the inability of these
methods to handle larger values of $d$. After closer examination of the
classification studies performed in the authors' original papers, we noticed
that classification accuracy estimates were reported only for small values of
$d$, thereby ignoring the inclusion of a larger number of variables. Hence, we
advise practioners to consider filtering a large number of features before
employing these methods.

@@

In our simulation study, we compared our proposed \emph{HDRDA} classifier to
several recently proposed classifiers, including the Dudoit, Pang, Tong, and Guo
classifiers, each of which assumes independence of the features conditional on
class membership. Despite the rapid computational performance of diagonal
classifiers and the reduction in the number of parameters to estimate,
\cite{Mai:2012bf} and \cite{Fan:2012iq} have noted that diagonal classifiers can
often yield inferior classification performance to other classification
methods. We have shown that the \emph{HDRDA} classifier often yields superior
classification accuracy to the diagonal classifiers considered, confirming the
assertions of \cite{Mai:2012bf} and \cite{Fan:2012iq}. Furthermore, the
implementation of our classifier in the {\tt sparsediscrim} R package was
comparable to that of the diagonal classifiers in terms of computational
runtime. Moreover, when preliminary variable selection was applied to the data
in our classification study, the time to perform model selection with our
proposed classifier was nearly instantaneous. Hence, our proposed \emph{HDRDA}
classifier is competitive in terms of classification accuracy for modern
small-sample, high-dimensional data sets and requires substantially less
computing time than standard regularized classifiers, such as the \emph{RDA}
classifier proposed by \cite{Friedman:1989tm}.
